---
title: "Covid 19-India"
date: 2020-08-23
author:  LiJin Joo [lijin.joo at gmail.com](mailto:lijin.joo@gmail.com) 
output: html_document
bibliography: covid19.bib
---

In this section, I will reproduce the estimates for indian COVID 19 cases between March 19 and May 17 in [@Basu2020]. You may find all necessary informatioin at [@Umich_covid]. 

```{r, include=FALSE}
options(warn=-1)
library(EpiEstim)
library(ggplot2)
library("gridExtra")
library(magrittr)
```

Down laoding the data files raw_data 1-5 from [@covid19indiaorg2020tracker], we can prepare a series of incidence as follows.
```{r, cache=TRUE}
data1=read.csv("raw_data1.csv", header=T)
data2=read.csv("raw_data2.csv", header=T)
data3=read.csv("raw_data3.csv", header=T)
data4=read.csv("raw_data4.csv", header=T)
data5=read.csv("raw_data5.csv", header=T)

series1<-as.Date(data1[,3], format = "%d/%m/%y")
series2<-as.Date(data2[,3], format = "%d/%m/%y")
series3<-as.Date(data3[,3], format = "%d/%m/%y")
series4<-as.Date(data4[,3], format = "%d/%m/%y")
series5<-as.Date(data5[,3], format = "%d/%m/%y")

all_cases <- c(series1, series2, series3, series4, series5)
cases_covid19_india<-as.numeric(table(all_cases))
dates_covid19_india <- names(table(all_cases))

head(cases_covid19_india, 10)
head(dates_covid19_india, 10)


```

From their supplment, they explained how they obtained the distribution of serial intervals as follows: 

> We use the “parametric_SI” estimation method and a 5-day window (“estimate_R” function, which was used to
> describe the progression of the outbreak in Wuhan). We also use a gamma distribution prior with a mean of 7 
> days and a standard deviation of 4.5 days, based on research by Wu and colleagues, for the generation time 
> (a distribution of the onset of disease used to estimate R).
> 

We estimate Rt for Covid19 of India as follows.

```{r, cache=TRUE}

    t_start <- seq(2, length(cases_covid19_india) - 4)
    t_end   <- t_start + 4
    
    Rt_covid19 <- EpiEstim::estimate_R(incid = cases_covid19_india, method = "parametric_si",
                     config = make_config(list(mean_si = 7, std_si = 4.5, si_parametric_distr = "G",
                            t_start = t_start, t_end = t_end, seed = 123)))
    
```

We see that the estimates are very close to the numbers from [@Basu2020] below.

> "The estimated effective reproduction number R for India was 3.36 (95% confidence interval (CI):
>[3.03, 3.71]) on March 24, 
> whereas the average of estimates from May 25 - May 31 stands at 1.27 (95% CI: [1.26, 1.28]). "
> 


```{r, cache=TRUE}

 tibble::tibble(
        date_num = Rt_covid19$dates
    ) %>% dplyr::left_join(
        Rt_covid19$R, by = c("date_num" = "t_end")
    ) %>%
        dplyr::select(
            date_num, t_start, 'Mean(R)', 'Quantile.0.025(R)', 'Quantile.0.975(R)'
        ) %>%
        tibble::add_column(date = dates_covid19_india) %>%
        dplyr::select(-date_num) %>%
        dplyr::select(date, tidyselect::everything()) %>%
        dplyr::slice(c(24:27, 87:94))

win <- 89
R_median <- Rt_covid19$R$`Median(R)`[win]
R_CrI <- c(Rt_covid19$R$`Quantile.0.025(R)`[win], Rt_covid19$R$`Quantile.0.975(R)`[win])
set.seed(2019) 
R_sample <- sample_posterior_R(Rt_covid19, n = 10000, window = win)

hist(R_sample, col = "grey", breaks=100, main = "R sampled from May 25 - 31")
abline(v = R_median, col = "red") 
abline(v = R_CrI, col = "red", lty = 2) 

```

The three graphs show the trends of COVID19 cases and Rt as well as SI density. 

```{r, cache=TRUE, fig2, fig.height = 3.5, fig.width = 10}
library(incidence)
case_covid19 <- plot(as.incidence(cases_covid19_india)) + ggtitle("Covid19-India-Cases")
rt_covid19 <- plot(Rt_covid19, "R") + theme(legend.position = "none") + ggtitle("Covid19-India-Rt")
si_covid19 <- plot(Rt_covid19, "SI") + theme(legend.position = "none") + ylab("SI frequency") + ggtitle("Covid19-India-SI distribution")

grid.arrange(case_covid19, rt_covid19, si_covid19, nrow = 1)

```





### Reference