---
title: "Evaluate {EpiEstim} on five epidemic outbreaks"
date: 2020-08-23
author: LiJin Joo [lijin.joo at gmail.com](mailto:lijin.joo@gmail.com) 
output: html_document
bibliography: covid19.bib
---

[@cori2013new] used the following five examples to demonstrate \{EpiEstim\} in thier paper. As the first step of \{EpiEstim\} evalution, we reproduced Figure1 and generated a table summarize the comparison of the current version and the result resported in Cori et al.  

```{r, include=FALSE}
options(warn=-1)
```
```{r}
library(EpiEstim)
library(ggplot2)

data(Measles1861)
data(Flu1918)
data(Smallpox1972)
data(SARS2003)
data(Flu2009) 
```
Each data containis at least two varriables, a serise of number of cases (i.e. incidence) and a distirbution of serial intervals (i.e. si_distr) as in Flu 2009. 

```{r}
head(Flu2009$incidence)
Flu2009$si_distr
```
 Several options are available for different estimators for a distribution of serial intervals in \{EpiEstim\}: 1) `nonparametric_si' with the supplied density weights (si_distr), 2) `parametric_si' with options in `contig', such as mean_si, std_si, or 3) learning a distribution of serial intervals from data or samples using mcmc through `make_contig' function. Details of each option can be learned from [@EpiEstim_user], [@EpiEstim_github], [@EpiEstim_R].

### Reproducing Figure 1 

With the provided si_distr, we can obtain Rt estimates non-parametrically with a default of 7 day sliding window as follows:
```{r, cache=TRUE, message=FALSE}
Rt_measles1861 <- estimate_R(incid = Measles1861$incidence,
                         method = "non_parametric_si",
                         config = make_config(list(si_distr = Measles1861$si_distr)))

round(head(Rt_measles1861$R[,c(1:4, 5, 8, 11)], 15), 1)

Rt_flu1918 <- estimate_R(incid = Flu1918$incidence,
                         method = "non_parametric_si",
                         config = make_config(list(si_distr = Flu1918$si_distr)))

round(head(Rt_flu1918$R[,c(1:4, 5, 8, 11)]),1)

Rt_Smallpox1972 <- estimate_R(incid = Smallpox1972$incidence,
                              method = "non_parametric_si",
                              config = make_config(list(si_distr = Smallpox1972$si_distr)))

round(head(Rt_Smallpox1972$R[,c(1:4, 5, 8, 11)], 10),1)

Rt_sars2003 <- estimate_R(incid = SARS2003$incidence,
                          method = "non_parametric_si",
                          config = make_config(list(si_distr = SARS2003$si_distr)))

round(head(Rt_sars2003$R[,c(1:4, 5, 8, 11)]), 1)

Rt_flu2009 <- estimate_R(incid = Flu2009$incidence$I,
                         method = "non_parametric_si",
                         config = make_config(list(si_distr = Flu2009$si_distr)))
round(head(Rt_flu2009$R[,c(1:4, 5, 8, 11)]), 1)


```

```{r, cache=TRUE, echo=FALSE, fig1, fig.height = 6.5, fig.width = 10}
library(incidence)
case_measles1861 <- plot(as.incidence(Measles1861$incidence)) + ggtitle("Measles 1861 ")
case_flu1918 <- plot(as.incidence(Flu1918$incidence)) + ggtitle("Flu 1918")
case_smallpox1972 <- plot(as.incidence(Smallpox1972$incidence)) + ggtitle("Smallpox 1972")
case_sars2003 <- plot(as.incidence(SARS2003$incidence)) + ggtitle("SARS 2003")
case_flu2009 <- plot(as.incidence(Flu2009$incidence$I)) + ggtitle("Flu 2009")

rt_measleas1861 <- plot(Rt_measles1861, "R") + theme(legend.position = "none", plot.title = element_blank())
rt_flu1918 <- plot(Rt_flu1918, "R") + theme(legend.position = "none", plot.title = element_blank())
rt_smallpox1972 <- plot(Rt_Smallpox1972 , "R")+ theme(legend.position = "none", plot.title = element_blank())
rt_sars2003 <- plot(Rt_sars2003, "R") + theme(legend.position = "none", plot.title = element_blank())
rt_flu2009 <- plot(Rt_flu2009, "R") + theme(legend.position = "none", plot.title = element_blank())

si_measleas1861 <- plot(Rt_measles1861, "SI") + theme(legend.position = "none", plot.title = element_blank()) + ylab("SI frequency")
si_flu1918 <- plot(Rt_flu1918, "SI") + theme(legend.position = "none", plot.title = element_blank()) + ylab("SI frequency")
si_smallpox1972 <- plot(Rt_Smallpox1972 , "SI")+ theme(legend.position = "none", plot.title = element_blank()) + ylab("SI frequency")
si_sars2003 <- plot(Rt_sars2003, "SI") + theme(legend.position = "none", plot.title = element_blank()) + ylab("SI frequency")
si_flu2009 <- plot(Rt_flu2009, "SI") + theme(legend.position = "none", plot.title = element_blank()) + ylab("SI frequency")


library("gridExtra")
grid.arrange(case_measles1861, case_flu1918 , case_smallpox1972, case_sars2003, case_flu2009,
             rt_measleas1861, rt_flu1918, rt_smallpox1972, rt_sars2003, rt_flu2009,
             si_measleas1861, si_flu1918, si_smallpox1972, si_sars2003, si_flu2009,
             ncol = 5, nrow = 3)

```

### Comparison 

Over the years, some features of \{EpiEstim\} have been updated thus minor differences are observed. Characteristics of each case curve and differences in the reproduced numbers are summarized as follows.

| Outbreak        | Cori et al (2013)                              | Reproduced   |
|-----------------|------------------------------------------------|----------------------------|
| `Measles 1861`  | initial : 4.3 (95 % CI: 2-8.2)    | 4.2 (95 Q: 1.8-8.3) at d=18  | 
| increased Rt in w3-4  | mid w3:  3.0 (95 % CI: 1.3–5.9)   | 2.8 (95 Q: 2.2-3.4) at d=39   |
| due to the high   | mid w4:  11.5 (95 % CI: 8.3–15.3)  | no second peak after d=29 last d Rt>10 | 
| transsibility        | w7: below 1   | 0.99 (95 Q: 0.7-1.3) at d=42       |
| `Flu 1918`      | end w2 : 1.4 (95 % CI: 1.0–1.9)|  1.4 (95 Q: 1.0-1.9) at d =8, 10 |
| 2 days of extremely| mid w5: 2.4 (95 % CI: 2.2–2.6) | 2.2 (95 Q: 2.4-2.6) at d = 31 |
| high incidence |   early w7: below 1  | 1.0 (95 Q: 0.9 - 1.0) at d = 42 |
| at d=31, 45    |                      | 0.9 (95 Q: 0.8-0.9) at d = 47 |
| `Smallpox 1972` | early w4: 3.4 (95% CI: 0.8–9.3)  | 3.4 (95 Q: 0.8-9.3) at d = 29 |
| a long induction | mid w6: 23.9 (95% CI: 19.0–29.5)  | 23.9 (95 Q: 19.1-29.5 at d = 38|
| period (4 wk) | early w8: below 1  | 0.9 (95 Q: 0.5-1.6) at d = 50 |
| `SARS 2003`     |  mid w3: 12.2 (95% CI: 10.0–14.7) | 12.2 (95 Q: 10.0-14.7) at d = 24 |
| 2 peaks | end w6: 2.6 (95% CI: 2.4–2.9) | 2.4 (95 Q: 2.6-2.9) at d = 41, 42 |
| | w7: below 1 |  0.9 (95 Q: 0.8-1.0) at d = 47 |
|`Flu 2009`       | early w2: 1.7 (95% CI: 1.0–2.6) | 1.7 (95 Q: 1.0–2.6) at d = 8| 
| a constant rate | end w2: 1.7 (95% CI: 1.2–2.2) | 1.7 (95 Q: 1.2–2.2) at d = 14 |
| | w4: 0.2 (95% CI: 0.1–0.5) | 0.2 (95 Q: 0.1–0.5) at d = 28 |
| | w5: 0.9 (95% CI: 0.3–2.0) | 0.9 (95 Q: 0.3–2.1) at d = 32 |
* Reproduced estimates are sample quantiles, median (quanteil at 0.025 - quantile at 0.975).

* 95Q indicates 95 sample quantile intervals. Cori et al. used a term of 95% CI but it seems they referring to the quantile intervals. 

* d indicates t_end in R_t result. 

### Conclusion

With fixed serial interval weights, the estimates are consistent in the reproducible studies.

A true challenge in a COVID 19 study is to obtain an sensible estimate for this distribution. 

### Reference

